\name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Model for a Precision Spectrograph

### 2020-05-21 paper/paper.tex
Megan Edits
including updated figures

\begin{abstract}\noindent%
\name\ is a hierarchical, non-parametric framework for wavelength calibrating spectrographs.  This method capitalizes on recent hardware improvements in highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities, more so than any parametric model.  We demonstrate the success of this method with data from the EXtreme PREcision Spectrograph, \expres, which includes a Menlo Systems laser frequency comb.  We find that \name\ returns more accurate wavelengths by greater than a factor of 2 over classic, non-hierarchical, polynomial solutions.  Employing \name\ wavelengths on \expres\ RV measurements of HD 3651 reduced the RMS of a planet fit by \lz{blah}.
\end{abstract}

@@

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season.  This suggests that such stabilized spectrographs vary along only a small number of axes in what you might call ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

@@

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven method to generate a wavelength model.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases primarily because the need for good wavelength calibration in this field is so great.  Irregardless, we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph.

@@

For each exposure, $n$, every line, $(\lambda,m)$, has an associated fitted detector position, $x(\lambda,m,n)$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a \code{NaN} (hardware not-a-number) for that echelle order instead.  Let there be $M$ lines per exposure.  \name\ reads in a $N \times M$ matrix of line positions for each of $M$ lines for each of $N$ exposures.

@@

We found that a cubic-spline interpolation that enforces monotonicicity, such as a Piecewise Cubic Hermite Interpolatin Polynoimal (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a continuous function, while the enforced monotonicity prevents large deviations that may befall a cubic spline.  Choice in interpolation scheme and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

 Flat-relative, optimally extracted LFC data is background-corrected using a univariate spline.  Each peak in an echelle order is then fit to a Gaussian.  The mean of the fitted Gaussian is taken to be the center of the line.  For each line, the ThAr wavelength solution is used to estimate the mode number of a line.  The precise wavelength is then calculated using
 \begin{equation}
 f_n = n \times  f_r + f_0
 \label{eq:lfc}
 \end{equation}
 where the repetition rate, $f_r$, is known from design of the LFC, and the offset frequency, $f_0$, has been determined experimentally.

@@

To assess the classic, polynomial-driven method of wavelength calibration, we take each LFC exposure and separate the lines into even and odd lines.  We then construct a wavelength solution using only the odd lines and use it to predict wavelengths for the even lines; i.e. a polynomial is fit to just the odd lines and then evaluated at the detector position of the even lines (see Equation \ref{eq:poly_wsol}).  We then generate a wavelength solution using only the even lines and use it to predict wavelengths for the odd lines.

@@

To test the denoising step of \name\ (\textsection \ref{sec:denoising}, \textsection\ref{sec:interp_time}), we employed \name\ on 90\% of all LFC exposures.  This means the basis vectors, $g_k(x,y,m)$,  and weights, $a_{nk}$, were constructed using only information from 90\% of all exposures.  We used the results to predict wavelengths for all the lines in the remaining 10\% of calibration exposures.  This allows us to test how well we can pinpoint the calibration state of the spectrograph using \name.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.png}
%\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.pdf}
\caption{Residuals for a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $/mps$.  Residuals to a classic, polynomial fit is shown in the top plot.  The center plot shows residuals to an interpolated wavelength solution.  The bottom plot shows residuals when a ThAr wavelength model generated by \name\ is interpolated onto LFC line positions.  The position of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure}

@@

This shows how the flexibility of an interpolated model can account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.  In the \expres\ data shown here, for example, the interpolated model accounted for \lz{detector effect STA will explain to us}.  This same flexibility should similarly allow interpolated wavelength solutions to account for position errors in pixel image blocks for different detectors.

Position errors in pixel image blocks.

@@

The wavelength solution could also move past interpolation.  For example, a Gaussian process could be used that is constrained to ensure monotonicity.  Replacing each step with a model will allow for full, hierarchical Bayesian inference.  This means the uncertainty from wavelength calibration could be completely marginalized out.  Doing so, however, is only worth it if the wavelength calibration is a significant fraction of the error budget.

@@

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  Residuals might then get worse before ultimately starting to get better again with large K, which marks te model over-fitting the data.  Though the returned RMS never gets worse, we find that the improvement plateaus between $K=6$ and $K=128$.  Comparisons of wavelengths returned by a $K=6$ model vs. a $K=32$ model show significant differences in less than 10 bluer lines, which are known to have greater error and variance in their measured line positions.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, additionally show most of the structure is being captured in the first six principal components with later components dominated by noise.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=0.56\textwidth]{Figures/pcA_lfc.png}
\includegraphics[width=0.42\textwidth]{Figures/pcAs_byDay.png}
\caption{Amplitude of the first two principal components show as a function of time (left) or fraction of a day (right) \lz{(Yeah, it's kind of a mess)}.  The top plots shows the amplitudes for the first principal component while the bottom plot shows the amplitudes for the second principal component.  Points mark the times at which we have calibration data.  Lines show the result of a linear interpolation.  In the right plot, the principal component amplitudes have been artificially offset by the median amplitude for each day so that all days are roughly on the same scale.  Here, points are colored by the MJD of each exposure.}
\label{fig:nightlyVariation}
\end{figure*}

@@

A simple linear interpolation would obviously give erroneously low values everywhere.  Due to the dispersion intrinsic to echelle spectrographs, the wavelength change between pixels grows greater with greater wavelengths.  This means that the function of wavelength vs. pixel across an echelle order will always be monotonically increasing and concave down everywhere.

@@

In terms of stability, \name\ should be adaptable to a wide range of instrument designs.  In truth, the amplitude of the variation matters little as long as all axes of variation can be captured by PCA or another dimensionality-reducing model.  Indeed, larger variations from less stabilized instruments might even make it easier for the PCA to identify these axes.

Most physical systems should have only a few dominant axis along which they vary, simply by what is possible from general mechanical engineering.  With PCA, this can be tested by plotting the power of the basis vectors describing the calibration space, which should fall quickly after a few vectors.  This only provides a measure of the variance in the PCA space, however, not an explicit test of the end-to-end variation in the resulting model.  This condition is therefore necessary but not sufficient if implementing \name with PCA.

@@

For the results presented here, the data was broken up into different eras of stability based on where the principal component amplitudes showed huge deviations.  This was done visually, though there exists many change-point detection algorithms that could be used.  There is a trade-off in including more exposures between introducing greater variation, but also more data to offer constraints that may be optimized.  Here, an era of stability was chosen  manually in order to focus on separating out time-domain intervals in which the data is relatively homogeneous, e.g. most exposures show the same lines.  Homogeneity is, of course, implicitly required when implementing PCA.

@@

Along those lines, we caution that with any wavelength solution, there is a perfect degeneracy between what is defined as the ``position of the line'' and the resultant wavelength solution.  If, for example, a cross correlation method is used to extract RVs from the data, a systematic difference may be introduced depending on the definition of what exactly is the line position, whether it be the center of a Gaussian fit, the edge of a complicated PSF, etc.  In principle, the best way to mitigate this uncertainty would be to use a calibration source that looks like a star.

@@

\name\ can be applied to any data that contains information about the calibration state of the spectrograph.  For example, though LFC and ThAr exposures are used as an example in this paper, \name\ would work similarly for an etalon or any other arc lamp with a list of lines and assigned wavelengths.  Modifications would have to be made with an etalon to include the free spectral range (FSR) as a free parameter that is fit for along with the low-dimensional basis.

Incorporating calibration lines taken through a simultaneous fiber would also be fairly simple, especially if the simultaneous reference is also used while taking calibration images.  The calibration lines through the simultaneous fiber could simply be concatenated to the array of calibration lines when the low-dimensional basis is constructed.  This not only allows the simultaneous calibration information to contribute to constructing the complete calibration space of the spectrograph, but also allows the simultaneous calibration information to uniquely pinpoint where the spectrograph is in that calibration space for any exposure, removing the need to interpolate the basis vector amplitudes from calibration exposures to science exposures.

@@

\acknowledgements{We thank the EXPRES team for their work on the instrument, software, and pipeline.  We thank Dan Forman-Mackey for illuminating discussion.  The data presented here made use of the Lowell Discovery Telescope at Lowell Observatory. Lowell is a private, non-profit institution dedicated to astrophysical research and public appreciation of astronomy and operates the LDT in partnership with Boston University, the University of Maryland, the University of Toledo, Northern Arizona University and Yale University. We gratefully acknowledge ongoing support for telescope time from Yale University, the Heising-Simons Foundation, and an anonymous donor in the Yale community. We especially thank the NSF for funding that allowed for precise wavelength calibration and software pipelines through NSF ATI-1509436 and NSF AST-1616086 and for the construction of EXPRES through MRI-1429365.  LLZ gratefully acknowledges support from the National Science Foundation Graduate Research Fellowship under Grant No. DGE1122492.}

>>>>

\begin{abstract}\noindent%
\name\ is a hierarchical, non-parametric framework for wavelength-calibrating spectrographs.  This method capitalizes on recent hardware improvements in \mb{highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.}  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities, more so than any parametric model.  We demonstrate the success of this method with data from the EXtreme PREcision Spectrograph, \expres, which includes a Menlo Systems laser frequency comb.  We find that \name\ returns more accurate wavelengths by greater than a factor of 2 over classic, non-hierarchical, polynomial solutions.  Employing \name\ wavelengths on \expres\ RV measurements of HD 3651 reduced the RMS of a planet fit by \lz{blah}.
\end{abstract}

@@

\mb{The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season.  This suggests that such stabilized spectrographs vary along only a small number of axes in what you might call ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.}

@@

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven method to generate a wavelength model.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases primarily because the need for good wavelength calibration in this field is so great.  Irregardless, we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph, though the precision of the returned wavelengths will depend on the available calibration sources (more discussion in Section \ref{sec:others} below) .

@@

For each exposure, $n$, every line, $(\lambda,m)$, has an associated fitted detector position, $x(\lambda,m,n)$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a \code{NaN} (hardware not-a-number) for that echelle order instead.  Let there be $P$ lines per exposure.  \name\ reads in a $N \times P$ matrix of line positions for each of $P$ lines for each of $N$ exposures.

@@

We found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a continuous function, while the enforced monotonicity prevents large deviations that may befall a cubic spline.  Choice in interpolation scheme and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

 Flat-relative, optimally extracted LFC data is background-corrected using a univariate spline.  Each peak in an echelle order is then fit with a Gaussian.  The mean of the fitted Gaussian is taken to be the center of the line.  For each line, the ThAr wavelength solution is used to estimate the mode number of a line.  The precise wavelength is then calculated using
 \begin{equation}
 f_n = n \times  f_r + f_0
 \label{eq:lfc}
 \end{equation}
 where the repetition rate, $f_r$, is known from design of the LFC, and the offset frequency, $f_0$, has been determined experimentally.

@@

To assess the classic, polynomial-driven method of wavelength calibration, we take each LFC exposure and separate the lines into even- and odd-indexed lines.  We then construct a wavelength solution using only the odd lines and use it to predict wavelengths for the even lines; i.e. a polynomial is fit to just the odd lines and then evaluated at the detector position of the even lines (see Equation \ref{eq:poly_wsol}).  We then generate a wavelength solution using only the even lines and use it to predict wavelengths for the odd lines.

@@

To test the denoising step of \name\ (\textsection \ref{sec:denoising}, \textsection\ref{sec:interp_time}), we employed \name\ on a randomly selected 90\% of all LFC exposures.  This means the basis vectors, $g_k(x,y,m)$,  and weights, $a_{nk}$, were constructed using only information from 90\% of all exposures.  We used the results to predict wavelengths for all the lines in the remaining 10\% of calibration exposures.  This allows us to test how well we can pinpoint the calibration state of the spectrograph using \name.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.png}
%\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.pdf}
\caption{Residuals for a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $\mps$.  Residuals to a classic, polynomial fit are shown in the top plot.  The center plot shows residuals to an interpolated wavelength solution.  The bottom plot shows residuals when a ThAr wavelength model generated by \name\ is interpolated onto LFC line positions.  The positions of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure}

@@

This shows how the flexibility of an interpolated model can account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.  In the \expres\ data shown here, for example, the interpolated model accounted for \lz{detector effect STA will explain to us}.  This same flexibility should similarly allow interpolated wavelength solutions to account for position errors in pixel image blocks for different detectors \citep{fischer2016, milakovic2020}.

@@

The wavelength solution could also move past interpolation.  For example, a Gaussian process could be used that is constrained to ensure monotonicity.  Replacing each step with a model will allow for full, hierarchical Bayesian inference.  This means the uncertainty from wavelength calibration could be completely marginalized out.  Doing so, however, is only worthwhile if the wavelength calibration is a significant fraction of the error budget.

@@

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  Residuals might then get worse before ultimately starting to get better again with large K, which marks the model over-fitting the data.  Though the returned RMS never gets worse, we find that the improvement plateaus between $K=6$ and $K=128$.  Comparisons of wavelengths returned by a $K=6$ model vs. a $K=32$ model show significant differences in less than 10 bluer lines, which are known to have greater error and variance in their measured line positions.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, additionally show most of the structure is being captured in the first six principal components with later components dominated by noise.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=0.56\textwidth]{Figures/pcA_lfc.png}
\includegraphics[width=0.42\textwidth]{Figures/pcAs_byDay.png}
\caption{Amplitude of the first two principal components shown as a function of time (left) or fraction of a day (right).  The top row of plots shows the amplitudes for the first principal component while the bottom row shows the amplitudes for the second principal component.  Points mark the times at which we have calibration data.  Lines show the result of a linear interpolation.  In the right plot, the principal component amplitudes have been artificially offset by the median amplitude for each day so that all days are roughly on the same scale.  Here, points are colored by the MJD of each exposure.}
\label{fig:nightlyVariation}
\end{figure*}

@@

A simple linear interpolation would give erroneously low values everywhere.  Due to the dispersion intrinsic to echelle spectrographs, the wavelength change between pixels grows greater with greater wavelengths.  This means that the function of wavelength vs. pixel across an echelle order will always be monotonically increasing and concave down everywhere.

@@

It is also possible to apply \name\ to etalon data with some modifications.  If the free spectral range (FSR) and therefor wavelength of each line is well known, then the etalon lines can be treated just like LFC lines with known wavelengths and orders.  Alternatively, the FSR can be included as a free parameter that is included when constructing the low-dimensional basis.  \lz{Hogg, I think more detail is needed here that I don't understand but would like to; can we discuss?}

In terms of stability, \name\ should be adaptable to a wide range of instrument designs.  \mb{In truth, the amplitude of the variation matters little as long as all axes of variation can be captured by PCA or another dimensionality-reducing model.  Indeed, larger variations from less stabilized instruments might even make it easier for the PCA to identify these axes. (Is this true?)}

Most physical systems should have only a few dominant axis along which they vary, simply by what is possible from general mechanical engineering.  With PCA, this can be tested by plotting the power of the basis vectors describing the calibration space, which should fall quickly after a few vectors.  This only provides a measure of the variance in the PCA space, however, not an explicit test of the end-to-end variation in the resulting model.  This condition is therefore necessary but not sufficient if implementing \name\ with PCA.

@@

For the results presented here, the data was broken up into different eras of stability based on where the principal component amplitudes showed huge deviations.  This was done visually, though there exists many change-point detection algorithms that could be used \citep{aminikhanghahi2017}.  There is a trade-off in including more exposures between introducing greater variation, but also more data to offer constraints that may be optimized.  Here, an era of stability was chosen  manually in order to focus on separating out time-domain intervals in which the data is relatively homogeneous, e.g. most exposures show the same lines.  Homogeneity is, of course, implicitly required when implementing PCA.

Simultaneous calibration data is easily incorporated, especially if the simultaneous reference is also used when taking calibration images.  The position of calibration lines taken through the simultaneous fiber can be concatenated to the array of line positions when the low-dimensional basis is constructed.  This not only allows the simultaneous calibration information to contribute to constructing the complete calibration space of the spectrograph, but also allows the simultaneous calibration information to uniquely pinpoint where the spectrograph is in that calibration space for any exposure, thereby removing the need to interpolate the basis vector amplitudes from calibration exposures to science exposures.

@@

\lz{Hogg: I'm still not sure I completely grasp the issue, can this be made clearer?}
Along those lines, we caution that with any wavelength solution, there is a perfect degeneracy between what is defined as the ``position of the line'' and the resultant wavelength solution.  If, for example, a cross correlation method is used to extract RVs from the data, a systematic difference may be introduced depending on the definition of what exactly is the line position, whether it be the center of a Gaussian fit, the edge of a complicated PSF, etc.  In principle, the best way to mitigate this uncertainty would be to use a calibration source that looks like a star.

@@

\name\ can be applied to any data that contains information about the calibration state of the spectrograph (see Section \ref{sec:others}).  For example, though LFC and ThAr exposures are used as an example in this paper, \name\ would work similarly for an etalon or any other arc lamp with a list of lines and assigned wavelengths.  Simultaneous calibration information can easily be accounted for by simply including the line position information from the simultaneous reference when constructing a low-dimensional basis of the instrument's calibration space.

@@

\acknowledgements{We thank the EXPRES team for their work on the instrument, software, and pipeline.  We thank Dan Foreman-Mackey for illuminating discussion.  The data presented here made use of the Lowell Discovery Telescope at Lowell Observatory. Lowell is a private, non-profit institution dedicated to astrophysical research and public appreciation of astronomy and operates the LDT in partnership with Boston University, the University of Maryland, the University of Toledo, Northern Arizona University and Yale University. We gratefully acknowledge ongoing support for telescope time from Yale University, the Heising-Simons Foundation, and an anonymous donor in the Yale community. We especially thank the NSF for funding that allowed for precise wavelength calibration and software pipelines through NSF ATI-1509436 and NSF AST-1616086 and for the construction of EXPRES through MRI-1429365.  LLZ gratefully acknowledges support from the National Science Foundation Graduate Research Fellowship under Grant No. DGE1122492.}
